---
title: "Project 7"
author: "bigMACs"
date: "2022-11-27"
output: html_document
bibliography: Bib_7.bib
---

```{r, echo= FALSE, include= FALSE}
library(ggplot2)
library(tidyverse)
library(vroom)
library(readr)
library(dplyr)
library(MuMIn)
#loading libraries

sub <- list.files(path = "Project 8 data", recursive = TRUE,
                  pattern = ".csv", 
                  full.names = TRUE)
#Reading all files in the class-wide Project 8 data folder and adding them to a list

dat.l <- list()
#making an empty list

for(i in sub){
  met.dat<- unlist(strsplit(i,"_"))
  group <- gsub("Project 8 data/", " ", met.dat[1])
  subject <- met.dat[2]
  ang <- as.numeric(met.dat[3])
  activity <- gsub(".csv","",met.dat[4])
  
  dat.l[[i]]<- read_delim(i, delim= " ", col_names = c("Reading","Force","Unit"))%>%
    mutate(Group=group,Subject=subject, Angle=ang, Activity = activity)
  
}
#Establishing meta-data and extracting information from the file names. Angles needed to be converted from character to numbers in order to calculate AIC values later and means/se of shifts.

dat <- do.call(rbind,dat.l) 
#Compiling the data into a single tibble named "dat"


dat.1 <- arrange(dat, Group, Subject, Angle) %>% 
  filter(Angle != 22.5)
#Reorganized the data by ascending order and got rid of an outlier at angle 22.5
#Group 7, subject 27, had a reading at angle 22.5, which is outside the range of angles we would like to analyze

dat.max_recording <- dat.1 %>% 
  group_by(Subject, Angle, Activity) %>% 
  summarize(max_recording = max(Force))
#Grouped data to find the maximum force of each reading for each person, angle, and treatment (control or fatigue)

dat.max_subject <- dat.1 %>% 
  group_by(Subject, Activity) %>% 
  summarize(max_subject = max(Force))
#To a different tibble, grouped data to find the maximum force for each individual under each activity (control or fatigue), regardless of angle


dat.max_joined <- dat.max_subject %>% 
  left_join(dat.max_recording) %>% 
  group_by(Subject, Angle, Activity) %>% 
  mutate(max_norm = max_recording/max_subject)
#Made a master dataset that contains both maximum subject force and maximum reading force, and calculated the normalized maximum forces for each reading. 
```

### Introduction

### Methods

### Results

#### Exploring Force Length Relationships

```{r, message = FALSE, echo = FALSE, fig.cap= "Figure 1: The normalized maximum forces produced by each individual under each condition (control or fatigue) across all tested angles between 45° to 157.5°" }
dat.max_joined %>% 
  ggplot(aes(Angle,max_norm,col=Activity))+geom_point()+ggtitle("Normalized Maximum Forces vs. Angle") + xlab("Angle (°)") + ylab("Normalized Maximum Forces") + labs(col = "Treatment")
```

As seen above,the relationship between force and angle appears to be quite noisy. This can be attributed to the sheer size of the data and the fact that not all subjects recorded forces at the same angles. However, a general trend can still be observed. The true maximum forces with a y-value of 1 appear at a smaller angle for the control treatment when compared to that for the fatigued treatment. 

<br>

```{r, echo= FALSE, include = FALSE}
dat.class_means <- dat.max_joined %>% 
  group_by(Angle, Activity) %>% 
  mutate(class_mean = mean(max_norm))
#Calculating means of normalized maximum forces by activity and angle to compare the force-length relationships in the class-wide data
```

```{r, message = FALSE, echo = FALSE, fig.cap= "Figure 2: The force length relationship displayed as mean normalized maximum forces produced at each angle and categorized by the treatment."}
dat.class_means %>% 
  ggplot(aes(x=Angle, y=class_mean, col = Activity))+geom_point()+ggtitle("Average Normalized Maximum Forces vs. Angle") + xlab("Angle (°)") + ylab("Mean Normalized Maximum Forces") + labs(col = "Treatment")
```
When we clean up the data by taking the sample mean of the normalized maximum forces, the trend is clearer. Under control conditions, the force-length relationship curve appears to peak at around 140°. Under fatigue conditions, the force-length relationship curve is still very noisy, because not not all subjects took their data at the same angles, some angle observations are only based on a couple of individuals. But it would appear that the peak occurs around 155°, just before a drop off at the end of the range. A typical force-length relationship would have the maximum output at an intermediate angle. For our study, we would expect that to be at 101.25°. However, both under control and fatigue treatments, our subjects produced a force-length relationship peak at a higher than expected angle. 

<br>

#### Force-Length Relationship Shift 

```{r, echo=FALSE, include=FALSE}
AICs <- dat.max_joined%>%
  group_by(Subject,Activity)%>%
  summarize(
    m2=AICc(lm(max_norm~poly(Angle,2))), #second order
    m3=AICc(lm(max_norm~poly(Angle,3))), #third order
    m4=AICc(lm(max_norm~poly(Angle,4))) #fourth order
  )%>%
  pivot_longer(m2:m4,names_to="model",values_to="AICc")

x.pred<-seq(45,157.5,length.out=1000)
fits <- dat.max_joined%>%
  group_by(Subject,Activity)%>%
  summarize(
    m2=predict(lm(max_norm~poly(Angle,2)),newdata=data.frame(Angle=x.pred)), #second order
    m3=predict(lm(max_norm~poly(Angle,3)),newdata=data.frame(Angle=x.pred)), #third order
    m4=predict(lm(max_norm~poly(Angle,4)),newdata=data.frame(Angle=x.pred)) #fourth order
  )%>%
  pivot_longer(m2:m4,names_to="model")%>%
  group_by(Subject,Activity,model)%>%
  summarize(theta_max=x.pred[which.max(value)])

best.models <- fits%>%
  left_join(AICs)%>%
  group_by(Subject,Activity)%>%
  mutate(best=AICc==min(AICc))%>%
  filter(best==TRUE)%>%
  dplyr::select(-best)
```

```{r, echo= FALSE, message= FALSE}
best.models%>%
  filter(Subject != 25) %>% 
  pivot_wider(id_cols=Subject,names_from = Activity,values_from=theta_max)%>%
  mutate(shift=fatigue-control)%>%
  ungroup()%>%
  summarise(mean.shift=mean(shift),se.shift=sd(shift)/sqrt(length(shift)))
```
Based on the best polynomial models, we calculated a mean shift of 8.65° in the maximum force producing angle between the fatigue treatment and the control treatment. This means there was an upwards shift in the force-length relationship when moving from control to fatigued muscles. Although the shift is sizeable, the standard error is also relatively high at 4.35°.

<br>

```{r, echo= FALSE, message= FALSE}
anova(lm(theta_max~Activity,best.models))
```
When we ran an ANOVA variance test between the maximum force producing angle and the treatment, we found that the realtionship, hence the shift in force-length relationship is statistically significant at the 5% level. 

### Discussion
The force-length relationship is a fundamental property of skeletal muscles. It signifies that the greatest force is outputted at its ideal/resting length, and the least force is generated when the muscle is stretched or shortened past this ideal length. This project explored the effect of length on muscle force generation, specifically the impact of fatigue on the FL relationship of the human upper limbs. Previous research has demonstrated that the FL relationship shifts promptly and long after eccentric fatigue (@yeung2008shift). 

Our class wide study found a mean shift of 8.65 degrees in the maximum force-producing angle between the fatigue and control treatments. These findings demonstrated an apparent upwards shift in the FL relationships when moving from the controlled to tired muscles. This shift was statistically significant at the 5% level, as confirmed through an ANOVA variance test. This data provides further analysis of human movement and muscle function, specifically upper arm biomechanics.

Further studies could be undergone exploring this relationship to find the effect of force and the length over a larger period of time. Studies could determine which angle/length fatigues muscles at a faster rate or even examine how these angles impact the recovery times of specific muscles. A previous study examined a similar relationship and found that the shift in the angle-force curve was proportional to the degrees of muscle damage. This study aimed to explore and compare the magnitude and time course of the change in the angle-force curves, both before and four consecutive days after eccentric and isometric exercise (@philippou2004changes). Further studies between fatigue and time within the FL relationship could point to an optimal angle/length and time to work a specific muscle, enhancing individuals' physical therapy and recovery times.

### Contributions

### References

